Sampling Strategies in Decisions from Experience

### 2022-03-24 manuscript.Rmd
Preliminary Draft Model

bibliography      : references_sampling-in-dfe.bib
csl               : /Users/Linus Hof/Documents/Psychologie/ARC/sampling-in-dfe/apa.csl
appendix          : manuscript_appendix.Rmd

@@

Following @kolmogorovFoundationsTheoryProbability1950 and adopting the symbolic notation from @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, let the tuple $(\Omega, \mathcal{F}, P)$ be a probability space.
Accordingly, $\Omega$ is a finite sample space with each element $\omega$ in $\Omega$ denoting a possible consequence of choosing the respective prospect.
$\mathcal{F}$ is the power set $\mathcal{P}(\Omega)$ with each element $A$ in $\mathcal{F}$ being a subset of $\Omega$, also denoted a random event.
$P$ is the probability measure
$$
P: \mathcal{F} \to [0,1]
\tag{1}
$$
assigning the elements of $\mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.
Now, the random variable is a function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{2}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ denoting a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the pushforward measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{3}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(\{\omega: X(\omega) = x \in A'\})
\; .
\tag{4}
$$
Note that the measure in (3) and (4), respectively, is a random variable's distribution and therefore a prospect. 
Accordingly, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
\{P_{X_1}, ... , P_{X_k}\}
\; ,
\tag{5}
$$
where each element $P_X$ can itself be considered a set of outcome-probability pairs $\{(x_i, p_X(x_i)\}$.

@@

The choices people actually make between prospects are often described in terms of deviations from the principle of mathematical expectation, according to which the prospect with the largest expected value (EV)
$$
EV = \sum_i^n x_i \times p_X(x_i)
\tag{6}
$$ 
should be chosen (EV maximization).
To *describe* how people's actual choices deviate from this principle [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations],^[<!--Open footnote-->@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects, e.g., the *certainty effect* and the *reflection effect*, which were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the principle of mathematical expectation and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].<!--Close footnote-->]
CPT and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing^[<!--Open footnote-->Note that CPT may be amended by a stochastic choice rule to account for the probabilistic nature of choices [@rieskampProbabilisticNaturePreferential2008]. Stochastic choice rules [@stottCumulativeProspectTheory2006, for an overview] make the choice of a prospect more probable the better its valuation relative to those of other prospects.<!--Close footnote-->]
the value
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{7}
$$ 
with objective outcomes $x_i$ being transformed by a value function $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P_X$.
More specifically, following @tverskyAdvancesProspectTheory1992, the objective outcomes are transformed by a value function
$$
\begin{aligned}
  v : \Omega' &\to \mathbb{R} \\
  x &\mapsto 
  \begin{cases}
     x_i^\alpha &\forall x_i \geq 0, \\
     -\lambda |x_i|^\alpha &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{8}
$$
with $\alpha \in [0,1]$ determining the degree of the function's concavity (convexity) over the positive (negative) outcome interval, and $\lambda > 1$ increasing the function's slope over the negative outcome interval only.
Each subjective value $v(x)$ is then multiplied (or: weighted) with a cumulative decision weight that takes the form
$$
\pi_i =
  \begin{cases}
     w^+(P(X \geq x_i)) - w^+(P(X > x_i)) \quad \forall x_i \geq 0, \\
     w^-(P(X \leq x_i)) - w^-(P(X < x_i)) \quad \text{else}
     \; , 
  \end{cases}
\tag{9}
$$
where $w$ is a monotonic increasing, nonlinear weighting function $w: [0,1] \to [0,1]$ satisfying $w^+(0) = w^-(0) = 0$ and $w^+(1) = w^-(1) = 1$.

@@

* The value of a cumulative decision weight depends on the estimated shape of the graph of $w$ or the weighting function's parameters, respectively.
That is, the shape of the graph of $w$ displays over which interval on the cumulative probability scale $[0,1]$ the images of $w$, $w(p)$, take values that are greater or smaller than the respective cumulative probabilities and how much greater or smaller these images are.
Hence, assuming a nonlinear graphical shape of $w$, the same numerical difference between objective cumulative probabilities may translate to cumulative decision weights of different value.

@@

The need to distinguish between the computational and the algorithmic level in the domain of decision making is exemplified by the fact that information about prospects may come in the form of descriptions or sampled outcomes.
Specifically, only rarely in peoples' daily life, the inputs to the decision-making process take the form of explicit descriptions of all outcome-probability pairs.
In such cases, people would make DfD, since the information from which the mind learns about the properties of prospects are complete descriptions thereof.
Yet, rather often, people make DfE, where the mind can learn about the latent properties of prospects only by experiental sampling over time [@hertwigDescriptionexperienceGapRisky2009, p. 517].
Accordingly, for the choice between $1, ..., k$ prospects, the inputs to the decision-making process are generated by a set of prospect-specific stochastic processes
$$
\{\{X_{1t}\}, ... , \{X_{kt}\}\}
\; ,
\tag{11}
$$
where each $\{X_{t}\}_{t \in \mathbb N}$ is a collection of independent and identically distributed (i.i.d.) random variables.^[Note that it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. However, in the sampling paradigm, i.i.d. random variables are commonly assumed.]
The inputs to DfE are then the realizations of the random variables in (11), which require different algorithms to be processed than do descriptions of the outcomes and their probabilities.

@@

Yet, there are distinct latent properties and combinations thereof that can be used to assess differences between prospects and for which evidence may be accumulated.
Respectively, this paper presents a computational model in which the latent properties, whose differences among prospects are represented in the decision variable, can change.
To this end, the model implements the summary and the round-wise decision strategy---for which either the prospect is chosen that produced the greater mean across all sampled outcomes or the prospect that won more comparisons between single sampled outcomes, respectively [see @hillsInformationSearchDecisions2010]---into a common evidence accumulation model.
That is, the model accumulates the evidence obtained from a sequence of comparisons between sample means, with the size of the samples underlying each mean comparison decreasing with increasing switching probabilities.
Accordingly, whereas the decision variables of the model capture how much (more) comparisons between sample means a prospect has won, the evidence accumulation process more closely resembles the round-wise strategy for high switching probabilities and small sample sizes, and the summary strategy for low switching probabilities and large sample sizes.
Now, since for a large overall number of sampled outcomes the summary strategy is more indicative for differences in the EV and the round-wise strategy is more indicative for differences in the likelihood to produce a higher outcome than the respective other prospect, the proposed model shows how the mind may rely on different sampling strategies to change for which differences in the latent properties of prospects evidence is accumulated.

@@

The model assumes that evidence is accumulated over a sequence of mean-based comparisons between the samples.
The results of each comparison is defined as a random variable
$$
\begin{aligned}
  Z : \mathbb{R} &\to \mathbb{N} \\
  (\overline{X} - \overline{Y}) &\mapsto 
  \begin{cases}
     1 &\text{if} \quad \overline{X} > \overline{Y}, \\
     -1 &\text{if} \quad \overline{X} < \overline{Y}, \\
     0 &\text{else}
     \; . 
  \end{cases}
\end{aligned}
\tag{}
$$
The evidence accumulation process can therefore be defined as a random walk 
$$
D_n = \sum_t^n Z_t \quad n \in \mathbb N
$$

@@

<!-- Pachur et al. ... -->

```{r}
predictions <- tibble("Simulation Pararameter" = c("s", "$\\theta$"),
                      "$\\gamma$" = c(0,0),
                      "$\\delta$" = c(0,0),
                      "$\\alpha$" = c(0,0))
apa_table(predictions, caption = "Model Predictions", escape = FALSE)
```

@@

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

```{r echo = FALSE, results = 'asis', cache = FALSE}
papaja::render_appendix('manuscript_appendix.Rmd')
```

>>>>

bibliography      : references_sampling-in-dfe.bib
csl               : apa.csl

@@

Following @kolmogorovFoundationsTheoryProbability1950 and adopting the symbolic notation from @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, let the tuple $(\Omega, \mathcal{F}, P)$ be a probability space.
Accordingly, $\Omega$ is a finite sample space with each element $\omega$ in $\Omega$ denoting a possible consequence of choosing the respective prospect.
$\mathcal{F}$ is the power set $\mathcal{P}(\Omega)$ with each element $A$ in $\mathcal{F}$ being a subset of $\Omega$, also denoted a random event.
$P$ is the probability measure
$$
P: \mathcal{F} \to [0,1]
\tag{1}
$$
assigning the elements of $\mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.
Now, the random variable is a function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{2}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ denoting a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the pushforward measure
$$
P' : \mathcal{F'} \to [0,1]
\tag{3}
$$
with 
$$
P'(A' \in \mathcal{F'}) := P(\{\omega: X(\omega) = x \in A'\})
\; .
\tag{4}
$$
Note that the measure in (3) and (4), respectively, is a random variable's distribution and therefore a prospect. 
Accordingly, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
\{P'^1, ... , P'^k\}
\; ,
\tag{5}
$$
where each element $P'$ can itself be considered a set of outcome-probability pairs $\{(x_i, p_X(x_i)\}$.

@@

The choices people actually make between prospects are often described in terms of deviations from the principle of mathematical expectation, according to which the prospect with the largest expected value (EV)
$$
EV = \sum_i^n x_i \times p_X(x_i)
\tag{6}
$$ 
should be chosen (EV maximization).
To *describe* how people's actual choices deviate from this principle [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations],^[<!--Open footnote-->@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects, e.g., the *certainty effect* and the *reflection effect*, which were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the principle of mathematical expectation and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].<!--Close footnote-->]
CPT and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing^[<!--Open footnote-->Note that CPT may be amended by a stochastic choice rule to account for the probabilistic nature of choices [@rieskampProbabilisticNaturePreferential2008]. Stochastic choice rules [@stottCumulativeProspectTheory2006, for an overview] make the choice of a prospect more probable the better its valuation relative to those of other prospects.<!--Close footnote-->]
the value
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{7}
$$ 
with objective outcomes $x_i$ being transformed by a value function $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P'$.
More specifically, following @tverskyAdvancesProspectTheory1992, the objective outcomes are transformed by a value function
$$
\begin{aligned}
  v : \Omega' &\to \mathbb{R} \\
  x &\mapsto 
  \begin{cases}
     x_i^\alpha &\forall x_i \geq 0, \\
     -\lambda |x_i|^\alpha &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{8}
$$
with $\alpha \in [0,1]$ determining the degree of the function's concavity (convexity) over the positive (negative) outcome interval, and $\lambda > 1$ increasing the function's slope over the negative outcome interval only.
Each subjective value $v(x)$ is then multiplied (or: weighted) with a cumulative decision weight that takes the form
$$
\pi_i =
  \begin{cases}
     w^+(P(X(\omega) \geq x_i)) - w^+(P(X(\omega) > x_i)) \quad \forall x_i \geq 0, \\
     w^-(P(X(\omega) \leq x_i)) - w^-(P(X(\omega) < x_i)) \quad \text{else}
     \; , 
  \end{cases}
\tag{9}
$$
where $w$ is a monotonic increasing, nonlinear weighting function $w: [0,1] \to [0,1]$ satisfying $w^+(0) = w^-(0) = 0$ and $w^+(1) = w^-(1) = 1$.

@@

* The value of a cumulative decision weight depends on the estimated shape of the graph of $w$ or the weighting function's parameters, respectively.
That is, the shape of the graph of $w$ displays over which interval on the cumulative probability scale $[0,1]$ the images of $w$ take values that are greater or smaller than the respective cumulative probabilities and how much greater or smaller these images are.
Hence, assuming a nonlinear graphical shape of $w$, the same numerical difference between objective cumulative probabilities may translate to cumulative decision weights of different value.

@@

The need to distinguish between the computational and the algorithmic level in the domain of decision making is exemplified by the fact that information about prospects may come in the form of descriptions or sampled outcomes.
Specifically, only rarely in peoples' daily life, the inputs to the decision-making process take the form of explicit descriptions of all outcome-probability pairs.
In such cases, people would make DfD, since the information from which the mind learns about the properties of prospects are complete descriptions thereof.
Yet, rather often, people make DfE, where the mind can learn about the latent properties of prospects only by experiental sampling over time [@hertwigDescriptionexperienceGapRisky2009, p. 517].
Accordingly, for the choice between $1, ..., k$ prospects, the inputs to the decision-making process are generated by a set of prospect-specific stochastic processes
$$
\{\{X^1_t\}, ... , \{X^k_t\}\}
\; ,
\tag{11}
$$
where each $\{X_{t}\}_{t \in \mathbb N}$ is a collection of independent and identically distributed (i.i.d.) random variables.^[Note that it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. However, in the sampling paradigm, i.i.d. random variables are commonly assumed.]
The inputs to DfE are then the realizations of the random variables in (11), which require different algorithms to be processed than do descriptions of the outcomes and their probabilities.

@@

Yet, there are distinct latent properties that can be used to assess differences between prospects.
Respectively, this paper presents a computational model which implements the summary and the round-wise decision strategy---for which either the prospect is chosen that produced the greater mean across all sampled outcomes or the prospect that won more comparisons between single sampled outcomes, respectively [see @hillsInformationSearchDecisions2010]---into a common evidence accumulation model.
That is, the model accumulates the evidence obtained from a sequence of comparisons between sample means, with the size of the samples underlying each mean comparison decreasing with increasing switching probabilities.
Accordingly, whereas the decision variables of the model capture how much (more) comparisons between sample means a prospect has won, the evidence accumulation process more closely resembles the round-wise strategy for high switching probabilities and small sample sizes, and the summary strategy for low switching probabilities and large sample sizes.
Now, since for a large overall number of sampled outcomes the summary strategy is more indicative for differences in the EV and the round-wise strategy is more indicative for differences in the likelihood to produce a higher outcome than the respective other prospect, the proposed model shows how the mind may rely on different sampling strategies to change for which differences in the latent properties of prospects evidence is accumulated.

@@

We consider the sampling paradigm and a choice between two prospects denoted by the distributions $P'^X$ and $P'^Y$ of the discrete random variables $X$ and $Y$, respectively.
The model assumes that the information-processing chain starts at random with a stochastic process on one of the two random variables.
Specifically, it is assumed that an agent starts with equal probability to sample from one of the prospects, say $P'^X$, and generates a sequence of random variables $X_1, X_2,...$, where the mean over the sequence is computed or updated with each new outcome that is sampled.
The respective stochastic process $\{S^X_n\}_{n \in \mathbb{N}}$ is defined by
$$
S^X_n = \frac{1}{n} \sum_{t = 1}^n X_t, \quad n \in \mathbb{N}
\; ,
\tag{12}
$$
and terminates as soon as an outcome is sampled from the other prospect.
The switching probability $\psi \in (0,1]$, which is adopted prior the start of the information-processing chain and is fixed throughout, controls the probability with which outcomes from different prospects are sampled in direct succession.
In other words, the switching probability is the probability with which the stochastic process on the current random variable stops and a new stochastic process on the other random variable starts with the subsequently sampled outcome.
To implement an evidence accumulation process, the model assumes that each time after a new stochastic process was started and terminated on both random variables, their respective values are compared, with the prospect underlying the stochastic process with the greater final value receiving a round-win.
Accordingly, for the prospect $P'^X$, this stage is modeled by the mapping
$$
\begin{aligned}
  Z^X : \mathbb{R} &\to \mathbb{N} \\
  S^X_n - S^Y_n &\mapsto 
  \begin{cases}
     1 &\text{if} \quad  S^X_n > S^Y_n, \\
     0 &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{13}
$$
where $Z^X$ is a random variable that takes only the two values $0$ and $1$, denoting a lost or won mean comparison, respectively. 
The model assumes that a round consisting of two stochastic processes---i.e., one from each prospect---and the subsequent comparison of their values is repeated until the sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold is adopted prior the start of the information-processing chain and is fixed throughout.
For $P'^X$, then, we obtain a sequence of independent comparisons $Z^X_1, Z^X_2, ...$ and model the respective evidence accumulation process as a random walk $\{D^X_m\}_{m \in \mathbb{N}}$ defined by 
$$
D^X_m = \sum_{i = 1}^m Z^X_i, \quad m \in \mathbb{N}
\;,
\tag{14}
$$
resembling a dynamic decision variable.
Note that since (13) describes a Bernoulli trial, the random walk in (14) is based on a Bernoulli process.

Finally, note that the described model assumes two random walks---i.e., one for each prospect---approaching the same threshold, which is the absolute number of mean comparisons a prospect must win in order to be chosen.
However, one may also define a single random walk, resembling a common dynamic decision variable, which approaches a positive threshold $\theta^+$ and a negative threshold $\theta^-$.
Therefore, (13) may be adapted such that $Z$ takes the value $-1$ instead of $0$, such that the threshold determines how many more mean comparisons a prospect must win in order to be chosen.

@@

```{r eval=FALSE}
predictions <- tibble("Simulation Pararameter" = c("s", "$\\theta$"),
                      "$\\gamma$" = c(0,0),
                      "$\\delta$" = c(0,0),
                      "$\\alpha$" = c(0,0))
apa_table(predictions, caption = "Model Predictions", escape = FALSE)
```

@@
